The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-09 kalman_walk_paper.tex
edited speciation rate envelope calculation. added figure of phenotypic divergence to the speciation example. edited discussion, and minor edits to other sections.

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

\newcommand{\calA}{\mathcal{S}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

@@

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
  \plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW. I don't think we need to write an anodyne -- the current sentence is true enough.}
However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
%Previously it has been suggested that any idealized study of evolution is incomplete 
%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

@@

We present an analytical description of the set of all linear biological systems with identical phenotypes 
\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

\section*{Gene networks as linear dynamical systems}
%
\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
%
%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
\paragraph{Systems theory}
Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $S(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.

More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
  \begin{align}
    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\calA_0)
  \end{align}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

@@

For instance, the phenotypic dynamics of a diploid organism are a function of its two haploid genomes $x$ and $y$: $P(x,y) = \left(x + y\right)/2$. Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(x \times y \right) \rightarrow z$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes $x$ and $y$ are both in $\mathcal{K}$, their diploid systems $P(x,y)$, and their recombinant systems $P(z,z)$, may not be in $\mathcal{K}$. This is because $\mathcal{K}$ is not mathematically required to be closed under averaging and recombination.

If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

\paragraph{Sexual reproduction and speciation}
A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.

  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

@@

\begin{alignat*}{2}
      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
\end{alignat*}

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
    Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}

    \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$. If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}

At any given time, there will be a range of regulatory coefficients present in the population
due to segregating genetic variants.
Over many generations, even if selective pressures do not change,
this range of networks will shift 
as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
How much variation do we expect to find within a population?
Is this range limited by available variation or kept in check by selection?
How fast will a population explore the space of equivalent networks?
In this section we explore informally a general model for this situation,
in which a population drifts stochastically near a set of equivalent, optimal systems.
We work with a population of effective size $N_e$.

Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype ($\calA_n$ above).
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

\paragraph{Offspring.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same qualitative conclusions.
\plri{but put this in the appendix?}

@@

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} \js{In progress, see paragraph above.}
By the arguments above, the means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma$. \js{I think you mean $\sigma^{2}$?}
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. \js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
These are depicted in figure \ref{fig:conceptual_fig}.

@@

  \jss{ Assume that selection acts directly against organisms whose regulatory coefficients are not on the optimum manifold. Further assume that selection constrains this diversity to be, on average, a distance of $m$ from optimum. We would expect $\sigma_S = m$. Furtheremore, two individuals within a population may both be on the optimum manifold, but at different points, on average of $\sigma_N$. Since individuals a distance of $\sigma_N$ apart will produce $F_2$s off of the manifold of $\mu = \frac{\sigma_N}{2}\sin2\theta$, recombination load may constrain $\sigma_N$ such that $\mu \leq m = \sigma_S$ (where $\theta$ is the slope of the manifold -- as $\theta$ approaches $0$ or $1$, $\sigma_N \rightarrow \infty$.). We can estimate $\sigma_N = \frac{2\sigma_S}{\sin2\theta}$.  
  }
%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Parameter estimates for real systems}

To translate these results into real predictions, we need to know
the strength of stabilizing selection on the phenotype,
and the amount (and structure) of heritable variation in the genotype.
These are known at best only roughly \citep{felsenstein1988phylogenies},
so we aim for order-of-magnitude estimates.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

\newcommand{\calA}{\mathcal{A}}
\newcommand{\calH}{\mathcal{H}}
\newcommand{\diag}{\text{diag}}
\newcommand{\1}{\mathbbm{1}}

\newcommand{\Sys}{\mathcal{S}}

@@

%%%%%%%%%%%%%%%%%%%%%%
%\begin{multicols}{2}
\section*{Introduction}
Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
  \plr{Saying "stalled" reflects negatively on other fields - rephrase?} \js{I don't think it reflects negatively. Even if it did, as long as it's a true statement it should be fine.}.
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
  \plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.} \js{deleted HW. I don't think we need to write an anodyne -- the current sentence is true enough.}
However, verbal theories are often insufficient, if not downright misleading \citep{servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
%Previously it has been suggested that any idealized study of evolution is incomplete 
%without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. 
%
%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.
%
%
%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

@@

We present an analytical description of the set of all linear biological systems with identical phenotypes 
\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".} \js{added the word ``linear.''}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all linear biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.
%
\section*{Gene networks as linear dynamical systems}
%
\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
%
%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
\paragraph{Systems theory}
Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
  \begin{align}
    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

@@

For instance, diploid system coefficients are the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} = \left(\Sys + \Sys'\right)/2$. Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(\Sys \times \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes $\Sys$ and $\Sys'$ are both in $\mathcal{K}$, their diploid and recombinant systems may not be in $\mathcal{K}$. This is because $\mathcal{K}$ is not required to be closed under averaging or recombination.

If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes. 
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

@@

%\begin{alignat*}{2}
%      A(2) &= \left[\begin{array}{cc} 2 & -1 \\[6pt] 5 & -2 \end{array}\right] &&\qquad A(2.01) = \left[\begin{array}{cc} 2 -\frac{1}{101} & -1 + \frac{1}{101} \\[6pt] 5 + \frac{1}{99} & -2 + \frac{1}{101} \end{array}\right] \\
%        A(2.1) &= \left[\begin{array}{cc} 2 - \frac{1}{11} & -1 + \frac{1}{11} \\[6pt] 5 + \frac{6}{55} & -2 + \frac{1}{11} \end{array}\right] &&\qquad A(2.5) = \left[\begin{array}{cc} 2 - \frac{1}{3} & -1 + \frac{1}{3} \\[6pt] 5 + \frac{2}{3} & -2 + \frac{1}{3} \end{array}\right]
%\end{alignat*}

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
    Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}
Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.

At any given time, there will be a range of regulatory coefficients present in the population
due to segregating genetic variants.
Over many generations, even if selective pressures do not change,
this range of networks will shift 
as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
%How much variation do we expect to find within a population?
%Is this range limited by available variation or kept in check by selection?
%How fast will a population explore the space of equivalent networks?
%In this section we explore informally a general model for this situation,
%in which a population drifts stochastically near a set of equivalent, optimal systems.
%We work with a population of effective size $N_e$.
\js{commented out questions here b/c we don't answer them directly -- what do you think?}

Suppose that a set $x$ of coefficients that determine a system (this is $\Sys$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

\paragraph{Offspring.}
\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same qualitative conclusions.
\plri{but put this in the appendix?}

@@

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
By the arguments above, the means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma$. 
\js{I think you mean $\sigma^{2}$?}
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
These are depicted in figure \ref{fig:conceptual_fig}.

@@

\paragraph{Speciation rates under neutrality}
%
\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
%In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
%\begin{align*}
%  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
%\end{align*}
%\plr{sentences don't begin with "And"}
%And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
%\plr{what is $T$?}
%\begin{align*}
%  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
%\end{align*}
%the population mean architecture will move at rate
%\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
%\begin{align*}
%  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
%\end{align*}
%where $T$ is time measured in generations. 
%USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
%Thus the squared difference of it's phenotype will be,
%\begin{align*}
%  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
%\end{align*}
%and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
%   \begin{align*}
%     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
%\end{align*}
% It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
%\begin{align*}
%  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
%  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
%\end{align*}
%and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE. 
%
For cases where system structures and $\gamma$ can be inferred,
specific speciation rates can be estimated.
For other situations we offer a back of the envelope calculation.

Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
$\sigma_S$ directly and $\sigma_N$ indirectly via recombination load.
Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
to be,
  \begin{align}
    \exp\left(-c^2 \frac{T}{N_e} \right)
  \end{align}
%
%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
%  \begin{align}
%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
%  \end{align}
